A data-driven model of nucleosynthesis with chemical tagging in a lower-dimensional latent space

### 2019-07-15 article/mcfa.tex
updates

% NEED: Final model with # of loads and # of clusters so we can:
% 		[ ] Update figure of loads (exp10-comparison)
% 		[ ] Make a clustering plot in abundances
% 		[ ] Check to see if any chemical clusters correspond to known clusters.
% 		[ ] Update discussion w.r.t. final model loads, etc.
% 		[ ] Update conclusion w.r.t. final model loads/clusters, etc, and missing data.
% 		[ ] Update abstract w.r.t. final model loads/clusters.
% 		[ ] Upload PDF to Galah survey website
%		[ ] Send email to Galah team for comment

@@

\author[0000-0002-0583-5918]{David Dowe}
\affiliation{Faculty of Information Technology, 
			 Monash University, 
			 Wellington Rd, Clayton 3800, Victoria, Australia}

\author{the GALAH team}

\begin{abstract}
Chemical tagging seeks to identify unique star formation sites from
present-day stellar abundances.
Previous techniques have treated each abundance dimension as being
statistically independent, despite theoretical expectations that
many elements can be produced by more than one nucleosynthetic process.
In this work we introduce a data-driven model of nucleosynthesis 
where a set of latent factors (e.g., nucleosynthetic yields) contribute
to all stars with different scores, and clustering (e.g., chemical tagging) 
is modelled by a mixture of multivariate gaussians
in a lower-dimensional latent space.
We use an exact method to simultaneously estimate the factor scores for
each star, the partial assignment of each star
to each cluster, and the latent factors common to all stars, even in the
presence of missing data entries.
We use an information-theoretic Bayesian principle to estimate the number of
latent factors and clusters.
Using the second \Galah\ data release we find that five latent factors are
preferred to explain $N =$ 1,072 stars with 18 chemical abundances. 
We identify the rapid- and slow-neutron capture 
processes, as well as latent factors consistent with Fe-peak and 
$\alpha$-element production, and another where only K and Zn dominate.
When we consider $N \sim$ 100,000 stars with missing abundances we find 
another seven factors, as well as 16 components in latent space.
Despite these components showing separation in chemistry that is explained
through different yield contributions, none show significant structure in 
their positions or motions. We argue that more data, and joint priors 
on cluster membership that are constrained by dynamical models, are necessary
to realise chemical tagging at a galactic-scale.
We release accompanying software that scales well with the available data,
allowing for model parameters to be optimised in seconds given a fixed number
of latent factors, components, and $\sim10^7$ abundance measurements.
\end{abstract}

@@

Gravitationally bound star clusters have been useful laboratories for
testing the limits and utility of chemical tagging. Although biases arise when
only considering star clusters that are still gravitationally bound, the chemical
homogeneity of open clusters provides an empirical measure of how similar stars
would need to be before they could be tagged as belonging to the same
star formation site \citep{Mitschang:2014}. However, there are analysis
issues in understanding how precisely those chemical abundances can be measured
\citep{Bovy:2016}, and how chemically similar stars can be that did not form 
together \citep[doppleg\"angers;][]{Ness:2018}.
If open clusters were truely chemically homogeneous then under idealistic 
assumptions our ability to chemically tag the Milky Way would depend primarily
on the precision with which we can measure those chemical abundances in stars. 
Data-driven approaches to modelling stellar spectra are
improving upon this precision \citep{Ness:2015,Ness:2018a,Ness:2018b,
Casey:2016,Casey:2017,Ho:2017b,Ho:2017a,Leung;Bovy:2018}, but more work is
needed: astronomers have not yet developed unbiased estimators of chemical
abundances that saturate the Cram\'er-Rao bound \citep{Cramer:1946,Rao:1945}.

%There are further chemical tagging obstacles to overcome once all the chemical 
%abundance information in a stellar spectrum can be extracted. Most notably is a 
Chemical tagging experiments require a catalogue of precise chemical abundance
measurements for a large number of stars,
where those chemical abundances trace different nucleosynthetic pathways.
This is the primary goal of the \Galah\ survey \citep{DeSilva:2015,Martell:2017,Buder:2018},
a stellar spectroscopic survey that uses the High Efficiency and Resolution 
Multi-Element Spectrograph \citep[\project{HERMES};][]{Sheinis:2015} on the Australian 
Astronomical Telescope (AAT).  \Galah\ will observe up to $10^6$ stars in the 
Milky Way, and measure up to 30 chemical abundances for each star. This includes
light odd-Z elements (e.g., Na, K), elements produced through
alpha-particle capture (e.g., Mg, Ca, Ti), and elements produced
through the slow (e.g., Ba) and rapid neutron-capture process
(e.g., Eu). No other spectroscopic survey provides an equivalent set of
chemical abundances for a comparable number of stars.

@@

Factor analysis is a common statistical approach for describing correlated 
observations with a lower number of latent variables \citep[e.g.,][]{Thompson:2004}.
Related techniques include principal component analysis \citep{Hotelling:1933} and its
variants \citep{Tipping;Bishop:1999}, singular value decomposition \citep{Golub:1970}, and other
matrix factorization methods. While factor analysis on its own is a useful
dimensionality reduction tool to identify latent factors that contribute to
the chemical abundances of stars \citep[e.g.,][]{Ting:2012,Price-Jones:2018}, factor
analysis cannot describe clustering in the data (or latent) space.
Similarly, clustering techniques applied to chemical abundances to date 
\citep[e.g.,][]{Hogg:2016} do not account for the lower effective dimensionality in
elemental abundances.

@@

\noindent{}where $\factorloads$ is a $\NumLatentFactors \times \NumDimensions$ 
matrix of factor loads that is common to all data points, $\NumLatentFactors$ is
the number of latent factors, and the factor scores 
for the $\numdata$th data point
\begin{equation}
	\factorscores_\numdata \sim \mathcal{N}(\vec\xi_\numcomponents, \vec\Omega_\numcomponents)
\end{equation}
\noindent{}are drawn from\footnote{For clarifying nomenclature across disciplines, the terminology $z \sim \mathcal{N}(0, 1)$ indicates that the $z$ variable \emph{is drawn from} a standard normal distribution.} the $\numcomponents$th multivariate normal distribution.
The mean vector $\vec\mu$ describes the mean datum in each dimension.
The factor scores for all data points $\factorscores$ is then a 
$\NumData \times \NumLatentFactors$ matrix, where each data point has a partial
association to the components in latent space. 
We assume $\vec{e} \sim \mathcal{N}\left(\vec{0}, \textrm{diag}(\specificvariance)\right)$
is independent of the latent space, and $\specificvariance$ is a
vector of variances in each $\NumDimensions$ abundance dimensions.
In this model each data point can be represented as being drawn
from a mixture of multivariate normal components, except the components
are \emph{clustered in the latent space} $\factorscores$ and projected
into the data space by the factor loads $\factorloads$.

We assume that the latent space is lower dimensionality than the
data space (e.g., $\NumLatentFactors < \NumDimensions$).
Within the context of stellar abundances, the factor loads
$\factorloads$ can be thought of as the \emph{mean} yields
of nucleosynthetic
events (e.g., $s$-process production from AGB stars averaged over
initial mass function and star formation history), and the
factor scores are analogous to the relative counts of those 
nucleosynthetic events. The clustering in factor scores
achieves the same as a clustering procedure in data space,
except we simultaneously estimate the latent processes that are
common to all stars (the so-called factor loads, analogous to 
nucleosynthetic yields). Within this framework a rare nucleosynthetic event
can still be described as a `factor load' $\factorloads_\numlatentfactors$, 
but its rarity would be represented by associated factor
scores being zero for most stars and thus have no contribution
to the observed abundances. In practice the factor loads can only be 
identified up to orthogonality and cannot be expressly interpreted as
nucleosynthetic yields because they have limited physical meaning
(we discuss this further in Section~\ref{sec:discussion}),
but this description of typical yields and relative event rates should
help build intuition for the model parameters, and provide context
within the astrophysical problem it is being applied.

@@

Without loss of generality the density of the mean-subtracted 
data $\vecdataunscaled - \vec\mu$ (which we hereafter will refer to simply as $\vecdata$) can be described as
\begin{equation}
	f(\vecdata; \vec\Psi) = \sum_{\numcomponents=1}^{\NumComponents}\weight_\numcomponents\phi(\vecdata;\factorloads\scoremeans_\numcomponents, \factorloads\scorecovs_\numcomponents\factorloads\transpose + \diag{\specificvariance})
	%\eye\specificvariance
\end{equation}
\noindent{}given $\NumLatentFactors$ common factor loadings and $\NumComponents$ components
clustered in the latent (factor score) space. Here the parameter
vector
$\vec\Psi$ includes $\{\factorloads,\vec\pi,\scoremeans,\scorecovs,\specificvariance\}$, and $\phi(\vecdata; \vec\theta)$
describes the density of a multivariate gaussian distribution,
and $\weight_\numcomponents$ describes the relative weighting of the $\numcomponents$th
component in latent space and $\sum\vec\weight^K = 1$.
The log likelihood is then given by
\begin{equation}
	\log\mathcal{L}(\vecdata|\vec\Psi) = \sum_{\numcomponents=1}^{\NumComponents}\log{f(\vecdata;\vec\Psi)} \quad . \label{eq:log-likelihood}
\end{equation}

@@

Here we describe how the model parameters are initialised.\footnote{This describes the default initialisation approach. Other approaches are available in the accompanying software.}
To initialise the factor loads $\factorloads$ we start by randomly drawing a $\NumDimensions \times \NumDimensions$ matrix from a Haar distribution \citep{Haar:1933},
which is uniform on the special orthogonal group $\textrm{SO}(n)$ and therefore guaranteed to return an orthogonal
matrix with a determinant of unity \citep{Stewart:1980}.
We denote the $\NumLatentFactors \times \NumDimensions$ left-most region of this
matrix to be $\mathbf{H}$, and by taking $\factorloads_\ast$ to be the Cholesky decomposition of $\mathbf{H}\transpose \mathbf{H}$, we initialise the factor loads as
\begin{eqnarray}
 	\factorloads = \mathbf{H}\left(\left(\factorloads_\ast\right)^{-1}\eye\right)
\end{eqnarray}
\noindent{}which ensures that $\factorloads$ is a $\NumLatentFactors \times \NumDimensions$
matrix of mutually orthogonal vectors.
%and that it is a valid rotation matrix
%$\factorloads\transpose \factorloads = \eye$.
We then initially assign each data point as belonging to one of the
$\NumComponents$ components using the \texttt{k-means++} algorithm \citep{Arthur:2007}
in the pseudo-latent space $\vecdata\factorloads$. Given the initial
factor loads and assignments, we then estimate the relative weights
$\vec\pi$, the mean factor scores of each component $\scoremeans$, and
the covariance matrix of factor scores of each component $\scorecovs$.
Finally, we initialise the specific variance $\specificvariance$ in each
dimension as the variance in each data dimension. Other initialisation 
methods for the latent factors include singular value decomposition \citep{Golub:1970}
or generating random noise with orthogonal constraints, and random assignment
is an alternative method that is available for initialising assignments.

@@

We use the expectation-maximization algorithm to estimate the model parameters
\citep{Dempster:1977}. With each expectation step we evaluate the log likelihood 
given the model parameters $\vec\Psi$ and we calculate the $\NumData \times \NumComponents$ responsibility 
matrix $\vec\tau$ whose entries are the posterior probability that the 
$\numdata$th data point is associated to the $\numcomponents$th component, given 
the data $\vecdata$ and the current estimate of the parameter vector $\vec\Psi$:
\begin{equation}
	\tau_{\numdata\numcomponents} = \frac{\weight_\numcomponents\phi(\vecdata_\numdata;\factorloads\scoremeans_\numcomponents, \factorloads\scorecovs_\numcomponents\factorloads\transpose + \diag{\specificvariance})}{\sum_{g=1}^{G}\weight_g\phi(\vecdata_\numdata;\factorloads\scoremeans_g, \factorloads\scorecovs_g\factorloads\transpose + \diag{\specificvariance})} \quad .
\end{equation}

At the maximization step we update our estimates of the parameters $\vec\Psi$,
conditioned on the data $\vecdata$ and the responsibility matrix $\vec\tau$.
The updated parameters estimates are found by setting the second derivative
of the log likelihood (Eq.~\ref{eq:log-likelihood}) to zero and solving for
the parameter values.\footnote{Strictly this introduces a statistical inconsistency in that we should update our parameter estimates by setting the second derivative of our information-theoretic objective function (Eq.~\ref{eq:message-length}) to zero instead of the log likelihood, but this inconsistency only becomes serious with small $N$ -- precisely the opposite situation of chemical tagging!}
In doing so this guarantees that every updated
estimate of the model parameters is guaranteed to increase the log likelihood.
Although there are no guarantees against converging on local minima, in 
practice it is sufficient to run expectation-maximization from multiple
initialisations (as we do) in order to ensure that the global minima is reached.
At the maximization step we first update our estimate of the relative weights 
$\vec\weight\nextstep$ given the responsibility matrix $\vec\tau$
\begin{equation}
	\weight_\numcomponents\nextstep = \frac{1}{\NumData} \sum_{\numdata=1}^{\NumData}\tau_{\numdata\numcomponents}
	% Include proof.
\end{equation}
\noindent{}where the $\vec{\Psi}\thisstep$ superscript refers to the current parameter estimates and $\vec{\Psi}\nextstep$ refers to the updated estimate for the next iteration.
The updated estimates of the mean factor scores 
$\scoremeans\nextstep$ for each component are then given by
\begin{eqnarray}
	\scoremeans_\numcomponents\nextstep = \scoremeans_\numcomponents\thisstep + \frac{\vec{G}\transpose(\vecdata\transpose - \factorloads\thisstep\scoremeans_\numcomponents\thisstep)\vec\tau_\numcomponents}{\NumData\weight_\numcomponents\nextstep}
\end{eqnarray}
\noindent{}where:
\begin{eqnarray}
	\vec{W} &=& (\scorecovs_\numcomponents\thisstep)^{-1}\eye \\
	\vec{V} &=& \left(\specificvariance\thisstep\right)^{-1} \\
	\vec{C} &=& (\vec{W} + (\factorloads\thisstep)\transpose\vec{V}\factorloads\thisstep)^{-1}\eye \\
	\vec{G} &=& \left[\vec{V} - \vec{V}\factorloads\thisstep\vec{C}\left(\vec{V}\factorloads\thisstep\right)\transpose\right]\factorloads\thisstep\scorecovs_k\thisstep \quad .
\end{eqnarray}

@@

\noindent{}where $\odot$ denotes is the entry-wise (Hadamard) product. 
%We repeat
%the expectation-maximization cycle for up to 10,000 steps or until the log likelihood
%improved by less than $10^{-5}$ between successive iterations. 
%Our software implementation allows for a positive regularisation term to be added
%along the diagonal of covariance matrices in latent space $\scorecovs$, for all experiments
%In all experiments 
Throughout this work we assume that the data are noiseless and
we do not add any observed errors to the constructed covariance matrices.
%Although our implementation
%allows for a positive regularisation term to be added along the diagonal of the covariance
%matrices in latent space $\scorecovs$, we set this regularisation term to zero for all
%experiments performed here. Similarly, we assume that the data are noiseless and we do
%not add any observed errors to the constructed covariance matrices.

@@

In this work we will assume that any missing data measurements are
missing at random. The missing data points can then be treated as unknown
parameters that must be solved for (and updated) at each iteration. Initially
we impute zeros for missing data entries in $\vec\data$, and at each iteration we update
these imputed valuee with our estimate of what the missing data values
are given the current model parameters. This ensures that the log-likelihood increases
with each iteration. Similarly, with each update we inflate
our estimates of the specific variances based on the fraction of missing
data points in each dimension
\begin{equation}
\specificvariance_d\nextstep = \specificvariance_d\nextstep\left(\frac{N}{N-M}\right)
\end{equation}
\noindent{}where $M$ is a the number of missing data entries in the $d$th dimension.
In Section \ref{sec:toy-model-missing-data} we show with a toy model that the latent
factor loads and scores can be reliably estimated even in the presence of high fractions
of missing data (e.g., 40\%), conditioned on our assumption that the data are missing
at random.

@@

%The Fisher information for $\factorloads$ \todo{needs to be double checked.}
\noindent{}Thus the message length to encode $\factorloads$ is given by
\begin{eqnarray}
	I(\factorloads) &=& -\log\left(\frac{p(\factorloads)}{\sqrt{|\mathcal{F}(\factorloads)|}}\right) \nonumber \\
I(\factorloads)	&=& \frac{1}{2}\textrm{Tr}(\textrm{Cov}(\factorloads\transpose)^{-1}\factorloads\transpose\factorloads) - \frac{1}{2}(\NumDimensions - \NumLatentFactors - 1)\log{|\factorloads\transpose\factorloads|} \nonumber \\
	&& \cdots \,\, + \frac{1}{2}\NumDimensions\NumLatentFactors\log{2} + \frac{1}{2}\NumDimensions\log{|\textrm{Cov(\factorloads)}|} - \Gamma\left(\frac{\NumDimensions}{2}\right) \nonumber \\ \label{eq:prior_L}%\nonumber \\
%	&& \cdots \todo{+\frac{1}{2}\log{|\mathcal{F}(\factorloads)|}} \quad . \label{eq:prior_L}
\end{eqnarray}

>>>>

@@

\author[0000-0002-0583-5918]{David L. Dowe}
\affiliation{Faculty of Information Technology, 
			 Monash University, 
			 Wellington Rd, Clayton 3800, Victoria, Australia}

\author[0000-0001-7516-4016]{Joss Bland-Hawthorn}
\affiliation{Sydney Institute for Astronomy, School of Physics,
			 A28, The University of Sydney, NSW 2006, Australia}
\affiliation{Center of Excellence for Astrophysics in Three Dimensions (ASTRO-3D),
			 Australia}
\affiliation{Miller Professor, Miller Institute, 
			 UC Berkeley, 
			 Berkeley, CA 94720, USA}

\author[0000-0002-4031-8553]{Sven Buder}
\affiliation{Research School of Astronomy and Astrophysics,
			 Australian National University,
			 Canberra, ACT 2611, Australia}

%\author{Janez Kos}
%\affiliation{Sydney Institute for Astronomy, School of Physics,
%			 A28, The University of Sydney, NSW 2006, Australia}

\author[0000-0003-3081-9319]{Geraint F. Lewis}
\affiliation{Sydney Institute for Astronomy, School of Physics,
			 A28, The University of Sydney, NSW 2006, Australia}

\author[0000-0002-3430-4163]{Sarah L. Martell}
\affiliation{School of Physics,
			 University of New South Wales, 
			 Sydney, NSW 2052, Australia}

\author[0000-0001-5344-8069]{Thomas Nordlander}			 
\affiliation{Research School of Astronomy and Astrophysics,
			 Australian National University,
			 Canberra, ACT 2611, Australia}
\affiliation{Center of Excellence for Astrophysics in Three Dimensions (ASTRO-3D), 
			 Australia}

\author[0000-0002-8165-2507]{Jeffrey D. Simpson}
\affiliation{School of Physics,
			 University of New South Wales, 
			 Sydney, NSW 2052, Australia}

\author[0000-0003-1124-8477]{Daniel B. Zucker}
\affiliation{Department of Physics and Astronomy,
			 Macquarie University, 
			 Sydney, NSW 2109, Australia}

\begin{abstract}
Chemical tagging seeks to identify unique star formation sites from
present-day stellar abundances.
Previous techniques have treated each abundance dimension as being
statistically independent, despite theoretical expectations that
many elements can be produced by more than one nucleosynthetic process.
In this work we introduce a data-driven model of nucleosynthesis 
where a set of latent factors (e.g., nucleosynthetic yields) contribute
to all stars with different scores, and clustering (e.g., chemical tagging) 
is modelled by a mixture of multivariate Gaussians
in a lower-dimensional latent space.
We use an exact method to simultaneously estimate the factor scores for
each star, the partial assignment of each star
to each cluster, and the latent factors common to all stars, even in the
presence of missing data entries.
We use an information-theoretic Bayesian principle to estimate the number of
latent factors and clusters.
Using the second \Galah\ data release we find that five latent factors are
preferred to explain $N =$ 1,072 stars with 18 chemical abundances. 
We identify the rapid- and slow-neutron capture 
processes, as well as latent factors consistent with Fe-peak and 
$\alpha$-element production, and another where only K and Zn dominate.
When we consider $N \sim$ 100,000 stars with missing abundances we find 
another seven factors, as well as 16 components in latent space.
Despite these components showing separation in chemistry that is explained
through different yield contributions, none show significant structure in 
their positions or motions. We argue that more data, and joint priors 
on cluster membership that are constrained by dynamical models, are necessary
to realise chemical tagging at a galactic-scale.
We release accompanying software that scales well with the available data,
allowing for model parameters to be optimised in seconds given a fixed number
of latent factors, components, and $\sim10^7$ abundance measurements.
\end{abstract}

@@

Gravitationally bound star clusters have been useful laboratories for
testing the limits and utility of chemical tagging. Although biases arise when
only considering star clusters that are still gravitationally bound, the chemical
homogeneity of open clusters provides an empirical measure of how similar stars
would need to be before they could be tagged as belonging to the same
star formation site \citep{Bland-Hawthorn:2010b,Bland-Hawthorn:2010a,Mitschang:2014}. However, there are analysis
issues in understanding how precisely those chemical abundances can be measured
\citep{Bovy:2016}, and how chemically similar stars can be that did not form 
together \citep[doppleg\"angers;][]{Ness:2018}.
If open clusters were truely chemically homogeneous then under idealistic 
assumptions our ability to chemically tag the Milky Way would depend primarily
on the precision with which we can measure those chemical abundances in stars. 
Data-driven approaches to modelling stellar spectra are
improving upon this precision \citep{Ness:2015,Ness:2018a,Ness:2018b,
Casey:2016,Casey:2017,Ho:2017b,Ho:2017a,Leung;Bovy:2018}, but more work is
needed: astronomers have not yet developed unbiased estimators of chemical
abundances that saturate the Cram\'er-Rao bound \citep{Cramer:1946,Rao:1945}.

%There are further chemical tagging obstacles to overcome once all the chemical 
%abundance information in a stellar spectrum can be extracted. Most notably is a 
Chemical tagging experiments require a catalogue of precise chemical abundance
measurements for a large number of stars,
where those chemical abundances trace different nucleosynthetic pathways.
This is the primary goal of the Galactic Archaeology with \project{HERMES} (\Galah) survey \citep{DeSilva:2015,Martell:2017,Buder:2018},
a stellar spectroscopic survey that uses the High Efficiency and Resolution 
Multi-Element Spectrograph \citep[\project{HERMES};][]{Sheinis:2015} on the 3.9~m Anglo-Australian 
Telescope (AAT).  \Galah\ will observe up to $10^6$ stars in the 
Milky Way, and measure up to 30 chemical abundances for each star \citep{Bland-Hawthorn:2016}. This includes
light odd-Z elements (e.g., Na, K), elements produced through
alpha-particle capture (e.g., Mg, Ca, Ti), and elements produced
through the slow (e.g., Ba) and rapid neutron-capture process
(e.g., Eu). No other current or planned spectroscopic survey provides an equivalent set of
chemical abundances for a comparable number of stars.

@@

Latent factor analysis is a common statistical approach for describing correlated 
observations with a lower number of latent variables \citep[e.g.,][]{Thompson:2004}.
Related techniques include principal component analysis \citep{Hotelling:1933} and its
variants \citep{Tipping;Bishop:1999}, singular value decomposition \citep{Golub:1970}, and other
matrix factorization methods. While factor analysis on its own is a useful
dimensionality reduction tool to identify latent factors that contribute to
the chemical abundances of stars \citep[e.g.,][]{Ting:2012,Price-Jones:2018}, factor
analysis cannot describe clustering in the data (or latent) space. As a result,
some works have performed clustering and then required different latent factors for each (totally assigned) component \citep[e.g.,][]{EdwardsDowe1998}.
Similarly, clustering techniques applied to chemical abundances to date 
\citep[e.g.,][]{Hogg:2016} do not account for the lower effective dimensionality in
elemental abundances.

@@

\noindent{}where $\factorloads$ is a $\NumLatentFactors \times \NumDimensions$ 
matrix of factor loads that is common to all data points, $\NumLatentFactors$ is
the number of latent factors, and the factor scores 
for the $\numdata^\mathrm{th}$ data point
\begin{equation}
	\factorscores_\numdata \sim \mathcal{N}(\vec\xi_\numcomponents, \vec\Omega_\numcomponents)
\end{equation}
\noindent{}are drawn from\footnote{For clarifying nomenclature across disciplines, the terminology $z \sim \mathcal{N}(0, 1)$ indicates that the $z$ variable \emph{is drawn from} a standard normal distribution.} the $\numcomponents^\mathrm{th}$ multivariate
normal distribution, where each $\numcomponents^\mathrm{th}$ component has a $\NumLatentFactors$-dimensional mean and a dense $\NumLatentFactors \times \NumLatentFactors$ covariance matrix.
The mean vector $\vec\mu$ describes the mean datum in each dimension.
The factor scores for all data points $\factorscores$ is then a 
$\NumData \times \NumLatentFactors$ matrix, where each data point has a partial
association to the components in latent space. 
We assume $\vec{e} \sim \mathcal{N}\left(\vec{0}, \textrm{diag}(\specificvariance)\right)$
is independent of the latent space, and $\specificvariance$ is a
vector of variances in each $\NumDimensions$ abundance dimensions.
In this model each data point can be represented as being drawn
from a mixture of multivariate normal components, except the components
are \emph{clustered in the latent space} $\factorscores$ and projected
into the data space by the factor loads $\factorloads$. In a sense we
are using latent factor analysis as a form of dimensionality reduction and
simultaneously performing clustering in latent space.

We assume that the latent space is lower dimensionality than the
data space (i.e., $\NumLatentFactors < \NumDimensions$).
Within the context of stellar abundances, the factor loads
$\factorloads$ can be thought of as the \emph{mean} yields
of nucleosynthetic
events (e.g., $s$-process production from AGB stars averaged over
initial mass function and star formation history), and the
factor scores are analogous to the relative counts of those 
nucleosynthetic events. The clustering in factor scores
achieves the same as a clustering procedure in data space,
except we simultaneously estimate the latent processes that are
common to all stars (the so-called factor loads, analogous to 
nucleosynthetic yields). Within this framework a rare nucleosynthetic event
can still be described as a `factor load' $\factorloads_\numlatentfactors$, 
but its rarity would be represented by associated factor
scores being zero for most stars and thus have no contribution
to the observed abundances. In practice the factor loads can only be 
identified up to orthogonality and cannot be expressly interpreted as
nucleosynthetic yields because they have limited physical meaning
(we discuss this further in Section~\ref{sec:discussion}),
but this description of typical yields and relative event rates should
help build intuition for the model parameters, and provide context
within the astrophysical problem it is being applied.

@@

Without loss of generality the density of the mean-subtracted 
data $\vecdataunscaled - \vec\mu$ (which we hereafter will refer to simply as $\vecdata$) can be described as \citep[Sec. 5.2.9, p236;][]{Wallace:2005}
\begin{equation}
	f(\vecdata; \vec\Psi) = \sum_{\numcomponents=1}^{\NumComponents}\weight_\numcomponents\phi(\vecdata;\factorloads\scoremeans_\numcomponents, \factorloads\scorecovs_\numcomponents\factorloads\transpose + \diag{\specificvariance})
	%\eye\specificvariance
\end{equation}
\noindent{}given $\NumLatentFactors$ common factor loadings and $\NumComponents$ components
clustered in the latent (factor score) space. Here the parameter
vector
$\vec\Psi$ includes $\{\factorloads,\vec\pi,\scoremeans,\scorecovs,\specificvariance\}$, and $\phi(\vecdata; \vec\theta)$
describes the density of a multivariate Gaussian distribution,
and $\weight_\numcomponents$ describes the relative weighting of the $\numcomponents^\mathrm{th}$
component in latent space and $\sum\vec\weight_{\numcomponents} = 1$.
The log likelihood is then given by
\begin{equation}
	\log\mathcal{L}(\vecdata|\vec\Psi) = \sum_{\numdata=1}^{\NumData}\log{f(\vecdata;\vec\Psi)} \quad . \label{eq:log-likelihood}
\end{equation}

@@

Here we describe how the model parameters are initialised.\footnote{This describes the default initialisation approach. Other approaches are available in the accompanying software.}
To initialise the factor loads $\factorloads$ we start by randomly drawing a $\NumDimensions \times \NumDimensions$ matrix from a Haar distribution \citep{Haar:1933},
which is uniform on the special orthogonal group $\textrm{SO}(n)$ and therefore guaranteed to return an orthogonal
matrix with a determinant of unity \citep{Stewart:1980}.
We denote the $\NumLatentFactors \times \NumDimensions$ left-most region\footnote{The region choice is arbitrary. All that is required is that the randomly-generated matrix have mutually orthogonal vectors.}
 of this
matrix to be our initial guess of $\factorloads$, which provides a set of mutually
orthogonal vectors.
%$\mathbf{H}$, and by taking $\factorloads_\ast$ to be the Cholesky decomposition of $\mathbf{H}\transpose \mathbf{H}$, we initialise the factor loads as
%\begin{eqnarray}
% 	\factorloads = \mathbf{H}\left(\left(\factorloads_\ast\right)^{-1}\eye\right)
%\end{eqnarray}
%\noindent{}which ensures that $\factorloads$ is a $\NumLatentFactors \times \NumDimensions$
%#matrix of mutually orthogonal vectors.
%and that it is a valid rotation matrix
%$\factorloads\transpose \factorloads = \eye$.
We then initially assign each data point as belonging to one of the
$\NumComponents$ components using the \texttt{k-means++} algorithm \citep{Arthur:2007}
in the latent space. Given the initial
factor loads and assignments, we then estimate the relative weights
$\vec\pi$, the mean factor scores of each component $\scoremeans$, and
the covariance matrix of factor scores of each component $\scorecovs$.
Finally, we initialise the specific variance $\specificvariance$ in each
dimension as the variance in each data dimension. Other initialisation 
methods for the latent factors include singular value decomposition \citep{Golub:1970}
or generating random noise with orthogonal constraints, and random assignment
is an alternative method that is available for initialising assignments.

@@

We use the expectation-maximization algorithm to estimate the model parameters
\citep{Dempster:1977}. With each expectation step we evaluate the log likelihood 
given the model parameters\footnote{When evaluating the log likelihood we use the precision (sparse inverse) matrix of the Cholesky decomposition of the covariance matrix for computational efficiency and stability.} $\vec\Psi$, the message length, and the $\NumData \times \NumComponents$ responsibility 
matrix $\vec\tau$ whose entries are the posterior probability that the 
$\numdata$th data point is associated to the $\numcomponents$th component, given 
the data $\vecdata$ and the current estimate of the parameter vector $\vec\Psi$:
\begin{equation}
	\tau_{\numdata\numcomponents} = \frac{\weight_\numcomponents\phi(\vecdata_\numdata;\factorloads\scoremeans_\numcomponents, \factorloads\scorecovs_\numcomponents\factorloads\transpose + \diag{\specificvariance})}{\sum_{g=1}^{G}\weight_g\phi(\vecdata_\numdata;\factorloads\scoremeans_g, \factorloads\scorecovs_g\factorloads\transpose + \diag{\specificvariance})} \quad .
\end{equation}

At the maximization step we update our estimates of the parameters $\vec\Psi$,
conditioned on the data $\vecdata$ and the responsibility matrix $\vec\tau$.
The updated parameters estimates are found by setting the second derivative
of the log likelihood (Eq.~\ref{eq:log-likelihood}) to zero and solving for
the parameter values.\footnote{Strictly this introduces a statistical inconsistency in that we should update our parameter estimates by setting the second derivative of our information-theoretic objective function (Eq.~\ref{eq:message-length}) to zero instead of the log likelihood, but this inconsistency only becomes serious with small $N$ -- precisely the opposite situation of chemical tagging!}
In doing so this guarantees that every updated
estimate of the model parameters is guaranteed to increase the log likelihood.
Although there are no guarantees against converging on local minima, in 
practice it is sufficient to run expectation-maximization from multiple
initialisations (as we do) in order to ensure that the global minimum is reached.
At the maximization step we first update our estimate of the relative weights 
$\vec\weight\nextstep$ given the responsibility matrix $\vec\tau$
\begin{equation}
	\weight_\numcomponents\nextstep = \frac{1}{\NumData} \sum_{\numdata=1}^{\NumData}\tau_{\numdata\numcomponents}
	% Include proof.
\end{equation}
\noindent{}where the $\vec{\Psi}\thisstep$ superscript refers to the current parameter estimates and $\vec{\Psi}\nextstep$ refers to the updated estimate for the next iteration.
The updated estimates of the mean factor scores 
$\scoremeans\nextstep$ for each component are then given by
\begin{eqnarray}
	\scoremeans_\numcomponents\nextstep = \scoremeans_\numcomponents\thisstep + \frac{\vec{G}\transpose(\vecdata\transpose - \factorloads\thisstep\scoremeans_\numcomponents\thisstep)\vec\tau_\numcomponents}{\NumData\weight_\numcomponents\nextstep}
\end{eqnarray}
\noindent{}where:
\begin{eqnarray}
	\vec{W} &=& (\scorecovs_\numcomponents\thisstep)^{-1} \\
	\vec{V} &=& \left(\specificvariance\thisstep\right)^{-1} \\
	\vec{C} &=& (\vec{W} + (\factorloads\thisstep)\transpose\vec{V}\factorloads\thisstep)^{-1} \\
	\vec{G} &=& \left[\vec{V} - \vec{V}\factorloads\thisstep\vec{C}\left(\vec{V}\factorloads\thisstep\right)\transpose\right]\factorloads\thisstep\scorecovs_k\thisstep \quad .
\end{eqnarray}

@@

\noindent{}where $\odot$ denotes is the entry-wise product. 
%We repeat
%the expectation-maximization cycle for up to 10,000 steps or until the log likelihood
%improved by less than $10^{-5}$ between successive iterations. 
%Our software implementation allows for a positive regularisation term to be added
%along the diagonal of covariance matrices in latent space $\scorecovs$, for all experiments
%In all experiments 
Throughout this work we assume that the data are noiseless and
we do not add any observed errors to the constructed covariance matrices.
%Although our implementation
%allows for a positive regularisation term to be added along the diagonal of the covariance
%matrices in latent space $\scorecovs$, we set this regularisation term to zero for all
%experiments performed here. Similarly, we assume that the data are noiseless and we do
%not add any observed errors to the constructed covariance matrices.

@@

In this work we will assume that any missing data measurements are
missing at random. The missing data points can then be treated as unknown
parameters that must be solved for (and updated) at each iteration. Initially
we impute zeros for missing data entries in $\vec\data$, and at each iteration we update
these imputed value with our estimate of what the missing data values
are given the current model parameters. This ensures that the log-likelihood increases
with each iteration. Similarly, with each update we inflate
our estimates of the specific variances based on the fraction of missing
data points in each dimension
\begin{equation}
\specificvariance_d\nextstep = \specificvariance_d\nextstep\left(\frac{N}{N-M}\right)
\end{equation}
\noindent{}where $M$ is a the number of missing data entries in the $d$th dimension.
In Section \ref{sec:toy-model-missing-data} we show with a toy model that the latent
factor loads and scores can be reliably estimated even in the presence of high fractions
of missing data (e.g., 40\%), conditioned on our assumption that the data are missing
at random.

@@

%The Fisher information for $\factorloads$ \todo{needs to be double checked.}
\begin{widetext}
\noindent{}Thus the message length to encode $\factorloads$ is given by
\begin{eqnarray}
%	I(\factorloads) &=& -\log\left(\frac{p(\factorloads)}{\sqrt{|\mathcal{F}(\factorloads)|}}\right) \nonumber \\
I(\factorloads)	&=& \frac{1}{2}\textrm{Tr}(\textrm{Cov}(\factorloads\transpose)^{-1}\factorloads\transpose\factorloads) - \frac{1}{2}(\NumDimensions - \NumLatentFactors - 1)\log{|\factorloads\transpose\factorloads|} + \frac{1}{2}\NumDimensions\NumLatentFactors\log{2} + \frac{1}{2}\NumDimensions\log{|\textrm{Cov(\factorloads)}|} - \Gamma\left(\frac{\NumDimensions}{2}\right) \quad . \label{eq:prior_L}%\nonumber \\
%	&& \cdots \todo{+\frac{1}{2}\log{|\mathcal{F}(\factorloads)|}} \quad . \label{eq:prior_L}
\end{eqnarray}
